Individual and Differential Harm in Redistricting

### 2025-07-30 paper/harm.qmd
cut lots

The academic study of gerrymandering has not reflected this diversity.
Nearly all research has been on partisan gerrymandering, and has centered on how districting plans treat political parties, as captured in the *seats-votes curve* or notions of *partisan symmetry* [@tufte1973pb; @grofman1983measures; @king1987democratic; @king1989representation; @gelman1994unified; @thomas2013estimating; @katz2020].
A large number of formulas for quantifying partisan gerrymanders have also been developed [@stephanopoulos2015; @mcdonald2015unfair; @warrington18].
The study of racial gerrymandering has been largely shaped by the legal framework for challenging such gerrymanders: the Voting Rights Act of 1965 (VRA).
This literature is characterized by a focus on identifying the voting preferences of racial groups through techniques such as ecological inference [@oloughlin1982id; @grofman1991identifying; @king1997solution; @greiner2006ecological].
Work on local redistricting is even more sparse, in part because of data limitations [but see @siegel2013educational; @abott2020large; @vargas2021racial].

The dominance of partisan frameworks limits redistricting analysis, even in purely partisan cases.
By focusing only on party effects, existing tools overlook how partisan lines also have an effect of relative voting power by race, wealth, age, and urbanicity.

This paper addresses these limitations with a new framework of *individual and differential harm*.
Our proposed framework focuses on measuring the concentration or dilution of voting power to favor or disfavor a particular group.
This avoids the pitfalls of party-centric approaches.

The harm framework combines two key ideas, further motivated in Section @sec-motivation.
First, it operates at the level of the individual voter and examines whether each voter is represented by the candidate of their choice.
This unifies gerrymandering analyses across contexts and reflects the goal that elected officials represent the preferences of the voting public [see also @brunell2010].

Second, the proposed framework considers *counterfactual* districting plans, not just a single enacted plan.
Many factors influence the performance and qualities of a districting plan.
Without a well-defined comparison, it is impossible to differentiate whether certain aspects of the system, like its partisan performance, are consequences of laws and political geography or result from improper manipulation of the system for political gain [@chen:rodd:13].
For example, in Massachusetts, it is currently impossible to draw even one reliable Republican seat, despite 32% of voters being Republican [@duch:18].
The 9-0 Democratic composition reflects the homogeneity of the state's political geography, not partisan gerrymandering.
Once a set of districting rules has been agreed on, what matters is not how a plan measures up to a fixed ideal, but how it performs in comparison to counterfactual plans.
Many metrics rely on *implicit* counterfactuals, which may be irrelevant or infeasible.
By making the comparison set explicit, we can better reflect legal and geographic realities, while making the interpretation of our measures clearer.

The combination of individual voter choices and counterfactual districting plans defines *individual harm*: informally, a voter is harmed if they would be represented by their candidate of choice under a counterfactual plan.
By examining how harm varies across partisan, racial, or ideological lines, we link it back to traditional gerrymandering.
Section @sec-motivation gives a simple example Section @sec-indiv-harm formalizes it.

A key feature of our framework is its clear separation of theoretical quantities of interest and empirical strategies for estimating them [@katz2022].
This distinction is also missed by many critics of the efficiency gap [@veomett2018efficiency; @deford2020symmetry].
The challenge of estimating harm and differential harm from observed data is discussed in Section @sec-estimating.
Recent advances in redistricting simulation, which can generate a large number of random districting plans that respect relevant legal constraints, are critical to this effort [@fifield2020mcmc; @deford2019; @carter2019; @mccartan2020].
These algorithmically-generated sets of districting plans can provide a realistic set of counterfactuals for our measures.

The harm framework has important similarities with the common-law approach to voting rights litigation.
To proceed with a legal challenge to an alleged gerrymander, a voter must demonstrate that their individual right to vote has been harmed by the enacted district boundaries [@1964reynolds; @1962baker; @issakarl98].
This requires proof that *but for* the alleged gerrymander, the voter would have been more likely to elect their chosen candidate [@2004vieth; @2018league; @2019lewis].
Indeed, a primary factor in the Supreme Court's ruling that partisan gerrymandering is not justiciable was that existing redistricting measures are not clearly tied to individual injury [@2017gill].

Our proposed harm measures improve on exististing racial and partisan gerrymandering analytical approaches.
The harm approach corresponds well to measuring the Gingles criteria, can directly use data from ecological inference, and handles complex cases like multi-minority coalitions [@stephanopoulos2014south]
Further, we explicitly model the counterfactuals and their feasibility, allowing us to improve on existing issues with other measures such as partisan bias [@king1987democratic], efficiency gap [@stephanopoulos2015], and declination [@warrington18].
A more detailed discussion of the connections between the proposed framework and existing practices for analyzing racial and partisan gerrymandering may be found in Section @sec-connections.

Harm and differential harm offer practical value across diverse contexts.
Section @sec-nj analyzes New Jersey congressional districts, showing how traditional metrics can misidentify both the magnitude and direction of partisan gerrymanders.
Section @sec-alabama uses differential harm to disentangle racial and partisan effects in Alabama's districting.
Section @sec-boston applies the framework to local redistricting, quantifying the benefits of race-aware boundaries for Black voters in Boston.
Section @sec-discussion concludes.

@@

Consider *Minissouri*, a stylized representation of Missouri which is shown in Figure @fig-minissouri.
Its total population is 180, and it is divided into three districts with population between 54 and 66.
Minissouri has two cities and a substantial rural population with a partisan breakdown of 55.6% R to 44.4% D.
Voters' preferences are assumed to be stable across elections, and turnout is 100%.
In addition to partisan affiliation (red or blue), each voter also belongs to a racial group (squares or circles).

@@

Suppose the redistricting plan in Figure @fig-minissouri(a) has been enacted.
Though Republicans win the popular vote, Democrats win two of three seats.
In contrast, @fig-minissouri(b) is more proportional, electing two Republicans and one Democrat.

Intuitively, plan (a) unfairly packs Republicans into a rural district.
We can identify which voters are harmed by the enactment of plan (a) over the alternative (b).
These voters are highlighted in Figure @fig-minissouri(c).
For example, in the upper leftmost precinct, there are three Republican square voters who are placed into a Democratic district in plan (a) but would have been placed into a Republican district in plan (b), and thus are harmed.
In contrast, Republican voters in the northeast city are not harmed, since they reside in a Democratic district under both plans.

After identifying which voters are harmed by plan (a) compared to counterfactual plan (b), it is natural to examine the overall composition of these harmed voters.
Of the 46 harmed voters, 44 are Republicans and only 2 are Democrats.
These represent 44% of all Republicans and 2.5% of all Democrats.
We call the gap between these two rates the *differential harm*, and propose it as a general way to account for the impact of a districting plan on particular groups.
Here, then the partisan differential harm of plan (a) is $-41.5\%$, meaning that the average Republican voter in Mininois is 41.5 percentage points less likely to be represented by a Republican than the average Democrat is to be represented by a Democrat, *compared to plan (b)*.

We need not limit our consideration to partisanship.
We can also tally harm by *race*, which in Minissouri is correlated with both geography and partisanship.
Plan (a) harms 13 circle voters and 33 square voters---a differential racial harm of $-20\%$.
This means square voters are 20 percentage points more likely to be harmed than circle voters.
While this number cannot establish whether the drawers of plan (a) intended to harm square voters, or aimed to maximize Democratic representation, it quantifies the impact of plan (a) on these voters versus the alternative plan (b), on the same scale as differential partisan harm.

@@

We formalize and generalize the concepts in Section @sec-motivation.
Let $\Voters$ be the set of individuals, or *voters*, in the region and time period.
Individuals vote for a candidate to represent districts, which are indexed by $\Distrs$.

Individuals vote in elections, each of which produces a specific outcome $v$ from the set $\Elecs$ of all possible election outcomes.
In an election, voters either vote for a candidate running in their district, or abstain from voting.
We write $v(i)$ to mean the candidate that individual $i$ votes for in election $v$.
The set of all individual vote choices (which may include the choice to abstain from voting) fully determines the specific election outcome.

@@

In an election $v$, the votes are tallied and the candidate with the most votes wins.
Using potential outcomes notation, we denote the winner of voter $i$'s district under districting plan $q$ and election $v$ as $\win_i(v,q)$.
This notation obscures the complex relationships and feedback effects between individuals, elections, and districting plans.
For example, different districting plans may lead to different candidates running, depending on district partisanship and incumbent residency.
The assumptions we make on these relationships will be made more explicit in Section @sec-estimating, where we discuss how to estimate our measures in practice.

@@

The key intuition from Section @sec-motivation is that to be harmed, a voter must have a better outcome in the counterfactual world where a different districting plan was in place.
Letting $\tilde q$ denote a counterfactual districting plan, we define the *harm* to an individual $i$ in election $v$ held under districting plan $q$ as

@@

A voter is harmed only if they are not represented by the candidate of their choice under $q$ ($\ind\{v(i) \neq \win_i(v, q)\}$) but they would have been represented by their candidate of choice under the counterfactual plan $\tilde q$ ($\ind\{v(i) = \win_i(v, \tilde q)\}$).

@@

Our definition of harm is limited in two regards.
First, it considers only individuals who vote: people who abstain from voting have experienced zero harm.
People who do not vote express no formal preference in electoral outcomes, even if they have informal preferences.
The natural theoretical extension of this definition is to then separate individuals' candidate preferences from their vote choice; the latter defines election winners, but the former provides a more-broad harm measure.
Naturally, implementing this extension in practice would require auxiliary data, such as a large-scale survey, since candidate preferences for non-voters cannot be directly inferred from election returns.

A second limitation is that $h_i(v,q,\tilde q)$ ignores any preferences a voter may have between the candidates that they did not vote for.
For example, a voter may consider candidates A and B nearly interchangeable, but may strongly dislike candidate C.
If the voter ultimately voted for candidate A, this harm measure makes no distinction between candidate B or candidate C winning the voter's district.
This reflects the simplistic nature of the cost function implicitly assumed in our proposed measure.
The problem could be remedied by replacing the indicator functions in the measure with other cost functions that consider ordinal preferences.
Ballot-level ranked-choice vote data, or candidate ideology scores, could be used to estimate such a cost function.
Here, for simplicity and to avoid making further assumptions, we focus on the indicator-based cost function.
Section @sec-boston further discusses the implications of this particular specification in the context of primary elections.

@@

The randomness in $V$ and $\tilde Q$ allows us to compare the enacted districting plan to many plausible counterfactual alternatives, and across a range of electoral environments.
At times we may wish to study on a particular election or a particular counterfactual plan.
Treating $V$ and $\tilde Q$ as random variables allows us to condition on particular values of each.
A particular specification of the distribution of $V$ and $\tilde Q$ is required only when *estimating* our harm measures in a particular application, a topic we return to in Section @sec-estimating.

@@

Despite being only a scalar quantity, $H(q)$ actually fully captures the harm created by the plan $q$.
We can zoom in to specific dimensions of the plan by conditioning on values of $V$, $\tilde Q$, and $I$, or by taking expectations.
Thus, for example, the harm for a single individual $i$ becomes $H(q)\mid I=i$, and the electorate's average harm in a particular election $v$ becomes
$$
    \E[H(q)\mid V=v] = \Pr[v(I)\neq\win_I(v, q) \text{ and }v(I)=\win_I(v, \tilde Q)],
$$
that is, the probability that a randomly selected voter in this election is misrepresented under the adopted plan but would not have been under a randomly selected counterfactual.

## Differential Harm and Fairness {#sec-diff-harm}

@@

Once we have defined a set of groups, we say that a plan $q$ is *fair* for groups $\Groups$ in comparison to a counterfactual $\tilde Q$ if $\E[H(q)\mid I\in G_k]$ is constant across groups $G_k\in\Groups$. Fairness is analogous to the notion of *error rate balance* in the algorithmic fairness literature, which seeks to equalize false positive and false negative rates across protected classes [@zafar2017fairness; @chouldechova2017fair].
There, the implicit counterfactual is a perfect predictor.
Compared to the perfect predictor, the actual decision process harms individuals whenever it makes a false positive or false negative error.
Here, there can be no perfect districting plan, and the counterfactual instead consists of a collection of plausible alternatives.

Fairness is a strict condition: real-life plans are highly unlikely to be exactly fair.
To quantify deviations from fairness, we introduce the concept of *differential harm*, defined as
$$
  \Delta_{G_k-G_l}H(q) \dfeq \E[H(q)\mid I\in G_k] - \E[H(q)\mid I\in G_l],
$$
for two groups $G_k$ and $G_l$.
So $\Delta_{G_k-G_l}H(q)$ will be positive if voters in group $G_k$ are harmed at a higher rate than voters in group $G_l$, indicating that plan $q$ favors $G_l$ over $G_k$.
Using our definition of fairness, $\Delta_{G_k-G_l}H(q)=0$ for all $G_k,G_l\in\Groups$ if and only if $q$ is fair.

@@

Estimating these quantities from available data is another matter, of course, and requires additional assumptions, as we describe in @sec-methods.

@@

To illustrate the connections in the preceding section and to demonstrate the value of the proposed measures in a real-world setting, we provide a brief analysis of congressional districting in the state of New Jersey.
Since 1995, New Jersey congressional redistricting has been handled by a 13-member commission, with 6 political appointees from each party and one independent chair with *de facto* tiebreaking authority.
For the 2020 redistricting cycle, each party submitted a proposed plan to John Wallace Jr, the chair.
Wallace ultimately selected the Democratic plan for enactment, "simply because in the last redistricting, the map was drawn by Republicans."[^7] Here, in contrast, we evaluate the two plans on the basis of existing gerrymandering metrics as well as through the proposed harm framework.

[^7]: See Matt Friedman, "Democrats prevail in New Jersey redistricting with map that could sacrifice Malinowski," *Politico,* December 22, 2021.
    <https://www.politico.com/news/2021/12/22/new-jersey-redistricting-map-malinowski-525983> (accessed July 3, 2022).

```{r fig-nj-maps, fig.width=6.25}
#| fig.cap: |
#|      Baseline vote patterns (a), two hypothetical gerrymanders (b) and (c),
#|      and one randomly sampled plan (d).
#|      County lines are overlaid in white.
include_graphics("figures/nj_maps.pdf")
```

Figure @fig-nj-maps shows the baseline Democratic vote share by precinct, along with the Democratic and Republican proposals.[^8]
Using an average of the 2016 presidential and 2018 senatorial elections as a baseline, the Democratic proposal yields 9 Democratic seats and 3 Republican seats, while the Republican proposal has 7 Democratic and 5 Republican seats.

[^8]: The enacted plan is sourced from [All About Redistricting](https://redistricting.lls.edu/state/new-jersey/?cycle=2020&level=Congress&startdate=2021-12-22), and the GOP plan was provided by Harrison Neely, executive director for the GOP delegation to the New Jersey Congressional Redistricting Committee, following an Open Public Records Act request.

We sample 10,000 districting plans across four independent runs of the SMC algorithm of @mccartan2020, as implemented in the R package `redist` [@redist], setting constraints to favor compact plans with district boundaries that tend to follow county lines.
One such plan is shown in Figure @fig-nj-maps(d).
For each of these sampled plans and the two gerrymanders, we estimate the average harm and differential partisan harm versus the simulated baseline,[^9] and calculate the expected number of Democratic seats, the efficiency gap, partisan bias, the mean-median score, and partisan declination.
The distribution of these measures is shown in Figure @fig-nj-meas.

[^9]: We choose to calculate $\Delta_{\dem-\rep}H(q)$, so that positive values correspond to a plan that favors Republicans.

We first examine the values of each metric on the two proposals alone.
The estimate of differential partisan harm aligns with the expected partisan skew of each plan, with a value of $-3.6\%$ for the Democratic proposal and $2.0\%$ for the Republican proposal.
These values are relatively directly interpretable.
For example, the estimated differential harm of $-3.6\%$ means that an average Democratic voter is 3.6 percentage points less likely to be misrepresented than their average Republican counter-partisan.
The efficiency gap, declination, and partisan bias measures agree with the differential harm estimates in this regard, though the former two are much more discrete.
The efficiency gap and partisan bias estimate give roughly equal scores in absolute value to the Democratic and Republican plans, in comparison to differential harm and declination, which identify the Democratic proposal as slightly more extreme.
While the mean-median measure identifies the Republican proposal as favoring Republicans, it also scores the Democratic proposal as favoring the Republicans, albeit by less.

@@

These values can be better contextualized in comparison to the distribution across the sampled plans.
The Democratic proposal is expected to produce an average of 8.55 Democratic seats under our election model, more than 79% of all sample plans.
The Republican proposal is an outlier, yielding fewer seats on average than every single sampled plan.
This pattern is repeated for all of the four common gerrymandering measures, where the Republican plan is a clear outlier (all $p=0.0002$, except $p=0.002$ for partisan bias) while the Democratic plan is not ($p=0.34$ for the efficiency gap, $0.50$ for declination, $0.88$ for partisan bias, and $0.17$ for the mean-median score).
However, both the partisan bias estimate and the mean-median score again misidentify the direction of the bias of the Republican plan, classifying it as an extreme Democratic outlier.

Figure @fig-nj-pairs in the appendix expands on Figure @fig-nj-meas to also show the correlations between these various measures.
Expected Democratic seats, the differential harm estimates, the efficiency gap, and declination are all moderately correlated with each other.
Similarly, partisan bias and the mean-median score are moderately correlated, but very weakly correlated with the expected number of Democratic seats.
Thus there exist many plans which give Democrats a higher-than-average share of the seats, but which have partisan bias and mean-median score favoring Republicans.
New Jersey is just one state, and while some of the correlations between measures here are found in other states, some are not.
Appendix @sec-app-corr records these correlations across simulated plans in eight states.
The most consistent correlation is between estimated differential partisan harm and the expected number of seats won by each party, which averages 0.79 in the eight states we study.
Seats are, of course, more discrete and cannot directly identify the local areas where gains and losses are derived.

@@

We next apply our proposed framework to @2022merrill, ongoing VRA litigation in which civil rights groups have argued that the existing Black majority district packs Black voters, and so a second Black opportunity-to-elect district should be drawn.[^10]
Measuring differential harm by both race and party will allow us to cleanly separate the racial and partisan effects of the Alabama districting plan, something existing approaches are unable to do.

[^10]: Relevant court filings are available at <https://redistricting.lls.edu/case/milligan-v-merrill/>.

As discussed in Section @sec-racial, plaintiffs had to satisfy the first *Gingles* criterion by submitting one or more remedial maps demonstrating that a new Black opportunity-to-elect district could be drawn while meeting all other redistricting requirements.
In *Merrill,* the plaintiff's expert, Moon Duchin, submitted four such remedial plans, one of which is shown in Figure @fig-al-maps(e).
These plans form a natural counterfactual $\tilde Q$ to use in analyzing the challenged districting plan (Figure @fig-al-maps(d)), especially since the federal district court ruled in favor of the plaintiffs and found that their remedial plans satisfied the first *Gingles* criterion.
In Appendix @sec-app-alsampling, we repeat the analysis performed here on a different counterfactual consisting of plans sampled with a redistricting simulation algorithm; the results are broadly similar.

@@

This analysis is slightly more complicated than in New Jersey, since in order to understand the racial distribution of harm we need to estimate voting preferences by race.
As is custom in VRA analyses, we use an ecological inference model.
The model, detailed in Appendix @sec-app-alei, jointly estimates two turnout and vote choice by race and precinct across seven statewide elections using a setup following @king1997solution.
For example, we model precinct $j$'s observed turnout $n_{jv}$ in election $v$ and baseline turnout rates $\turn_{jr}$ by race $r$ as \begin{align*}
    n_{jv} \mid \turn_{jr}, \beta^{(t)}_v
        &\sim \Binlog\qty(
        \overbrace{\vap_j,}^{\substack{\text{voting-age}\\\text{population}}}
        \hspace{0.5em}
        \logit(\turn_j^\top
        \overbrace{\vb p_j}^{\mathclap{\substack{\text{precinct racial}\\\text{proportions}}}}) +
        \underbrace{\beta^{(t)}_v}_{\mathllap{\text{election-specific turnout shift}}}) \\
    \logit(\turn_j) \mid \mu^{(t)}, \Sigma^{(t)}
        &\sim \mathcal{N}\underbrace{\qty(\mu^{(t)}, \Sigma^{(t)})}_{\mathrlap{
        \text{where to cluster the precinct turnout rates}}},
\end{align*} where $\Binlog$ indicates the Binomial distribution with the probability parameter specified on the logit scale.
We model Democratic support rates $\supp_{jr}$ by precinct and race similarly.

>>>>

The academic study of gerrymandering has not reflected this diversity.
Nearly all research has been on partisan gerrymandering, and has centered on how districting plans treat political parties, as captured in the *seats-votes curve* or notions of *partisan symmetry* [@tufte1973pb; @grofman1983measures; @king1987democratic; @king1989representation; @gelman1994unified; @thomas2013estimating; @katz2020].
A large number of formulas for quantifying partisan gerrymanders have also been developed [@stephanopoulos2015; @mcdonald2015unfair; @warrington18].
The study of racial gerrymandering has been largely shaped by the legal framework for challenging such gerrymanders: the Voting Rights Act of 1965 (VRA).
This literature focuses on identifying the voting preferences of racial groups through techniques such as ecological inference [@oloughlin1982id; @grofman1991identifying; @king1997solution; @greiner2006ecological].
Work on local redistricting is even more sparse, in part because of data limitations [but see @siegel2013educational; @abott2020large; @vargas2021racial].

The dominance of partisan frameworks limits redistricting analysis, even in purely partisan cases.
By focusing only on party effects, existing tools overlook how partisan lines also have an effect of relative voting power on axes such as race, wealth, and urbanicity.
This paper addresses these limitations with a new framework of *individual and differential harm*.
Our proposed framework focuses on measuring the concentration or dilution of voting power to favor or disfavor a particular group.
This avoids the pitfalls of party-centric approaches.

The harm framework combines two key ideas, further motivated in @sec-motivation.
First, it operates at the level of the individual voter and measures whether each voter is represented by the candidate of their choice.
This unifies gerrymandering analyses across contexts and reflects the goal that elected officials represent the preferences of the voting public [see also @brunell2010].

Second, the proposed framework considers *counterfactual* districting plans, not just a single enacted plan.
Many factors influence the qualities of a districting plan, and without a well-defined comparison, it is impossible to differentiate which qualities, like its partisan performance, are consequences of laws and political geography or result from improper manipulation of the system for political gain @chen:rodd:13.
For example, in Massachusetts in 2010, it was impossible to draw even one majority-Republican seat, despite 32% of voters preferring Republicans @duch:18.
Thus the 9-0 Democratic Congressional delegation reflects the homogeneity of the state's political geography, not partisan gerrymandering.
Many existing districting metrics rely on *implicit* counterfactuals, which may be irrelevant or infeasible, as we discuss in [Appendix @sec-connections].
By making the comparison set explicit, we can better reflect legal and geographic realities, while making the interpretation of our measures clearer.

The combination of individual voter choices and counterfactual districting plans defines *individual harm*: informally, a voter is harmed if they would be represented by their candidate of choice under a counterfactual plan.
Once measured for each voter, researchers can examine the distribution of harm by geography, partisanship, race, or other factors.
@sec-motivation gives a simple example of this idea; @sec-indiv-harm formalizes it.

A key feature of our framework is its clear separation of theoretical quantities of interest and empirical strategies for estimating them [@katz2022].
The methods section discusses the challenge of estimating harm and differential harm from observed data.
Recent advances in redistricting simulation, which can generate a large number of random districting plans that respect relevant legal constraints, are critical to this effort [@fifield2020mcmc; @deford2019; @carter2019; @mccartan2020].
These algorithmically-generated sets of districting plans can provide a realistic set of counterfactuals for our measures.

The harm framework has important similarities with the common-law approach to voting rights litigation.
To proceed with a legal challenge to an alleged gerrymander, a voter must demonstrate that their individual right to vote has been harmed by the enacted district boundaries [@1964reynolds; @1962baker; @issakarl98].
This requires proof that *but for* the alleged gerrymander, the voter would have been more likely to elect their chosen candidate [@2004vieth; @2018league; @2019lewis].
Indeed, a primary factor in the Supreme Court's ruling that partisan gerrymandering is not justiciable was that existing redistricting measures are not clearly tied to individual injury [@2017gill].
The harm measure also captures or generalizes the three *Gingles* preconditions for vote dilution litigation under the Voting Rights Act of 1965, as discussed further in [Appendix @sec-connections].

We demonstrate the practical value of the proposed framework in three diverse contexts.
@sec-nj analyzes New Jersey congressional districts, showing how traditional metrics can misidentify both the magnitude and direction of partisan gerrymanders.
@sec-alabama uses differential harm to disentangle racial and partisan effects in Alabama's districting.
@sec-boston applies the framework to local redistricting, quantifying the benefits of race-aware boundaries for Black voters in Boston.
@sec-discussion concludes.

@@

Consider *Minissouri*, a stylized representation of Missouri which is shown in @fig-minissouri.
Its total population is 180, and it is divided into three districts with population between 54 and 66.
Minissouri has two cities and a substantial rural population with a partisan breakdown of 55.6% R to 44.4% D.
Voters' preferences are assumed to be stable across elections, and turnout is 100%.
In addition to partisan affiliation (red or blue), each voter also belongs to a racial group (squares or circles).

@@

Suppose the redistricting plan in @fig-minissouri(a) has been enacted.
Though Republicans win the popular vote, Democrats win two of three seats.
In contrast, @fig-minissouri(b) is more proportional, electing two Republicans and one Democrat.

Compared to plan (b), plan (a) packs Republicans into a rural district.
@fig-minissouri highlights the voters who are harmed by the enactment of plan (a) over the alternative (b).
For example, in the upper leftmost precinct, there are three Republican square voters who are placed into a Democratic district in plan (a) but would have been placed into a Republican district in plan (b), and thus are harmed.
In contrast, Republican voters in the northeast city are not harmed, since they reside in a Democratic district under both plans.

After identifying which voters are harmed by plan (a) compared to counterfactual plan (b), it is natural to examine the overall composition of these harmed voters.
Of the 46 harmed voters, 44 are Republicans and only 2 are Democrats.
These represent $44\%$ of all Republicans and $2.5\%$ of all Democrats.
We call the gap between these two rates the *differential harm*, and propose it as a general way to account for the impact of a districting plan on particular groups.
Here, then the partisan differential harm of plan (a) is $-41.5\%$, meaning that the average Republican voter in Mininois is $41.5$pp less likely to be represented by a Republican than the average Democrat is to be represented by a Democrat, *compared to plan (b)*.

We need not limit our consideration to partisanship.
We can also tally harm by *race*, which in Minissouri is correlated with both geography and partisanship.
Plan (a) harms 13 circle voters and 33 square voters---a differential racial harm of $-20\%$.
This means square voters are $20$pp more likely to be harmed than circle voters.
While this comparison alone cannot establish whether the drawers of plan (a) intended to harm square voters versus maximize Democratic representation, it quantifies the impact of plan (a) on these voters versus the alternative plan (b), on the same scale as differential partisan harm.

@@

We formalize and generalize the concepts in @sec-motivation.
Let $\Voters$ be the set of individuals, or *voters*, in the region and time period.
Individuals vote for a candidate to represent districts, which are indexed by $\Distrs$.

Individuals vote in elections, each of which produces a specific outcome $v$ from the set $\Elecs$ of all possible election outcomes.
<!-- In an election, voters either vote for a candidate running in their district, or abstain from voting. -->
We write $v(i)$ to mean the candidate that individual $i$ votes for in election $v$.
The set of all individual vote choices (which may include the choice to abstain from voting) fully determines the specific election outcome.

@@

In an election $v$, the votes are tallied and the candidate with the most votes wins.
Using potential outcomes notation, we denote the winner of voter $i$'s district under districting plan $q$ and election $v$ as $\win_i(v,q)$.
This notation obscures the complex relationships and feedback effects between individuals, elections, and districting plans.
For example, different districting plans may lead to different candidates running, depending on district partisanship and incumbent residency.
The assumptions we make on these relationships are made explicit by the statistical models used to estimate the harm measure; these models may be as simple or complex as the researcher chooses, as we discuss in the methods section.

@@

The key intuition from @sec-motivation is that to be harmed, a voter must have a better outcome in the counterfactual world where a different districting plan was in place.
Letting $\tilde q$ denote a counterfactual districting plan, we define the *harm* to an individual $i$ in election $v$ held under districting plan $q$ as

@@

Harm requires both terms to be true: the first measuring if a voter is not represented by their chosen candidate under $q$, and the second term measuring if they would have been represented by their chosen candidate under the counterfactual plan $\tilde q$.

@@

@@

The randomness in $V$ and $\tilde Q$ allows us to compare the enacted districting plan to many plausible counterfactual alternatives, and across a range of electoral environments.
At times we may wish to study on a particular election or a particular counterfactual plan.
Treating $V$ and $\tilde Q$ as random variables allows us to condition on particular values of each.
A particular specification of the distribution of $V$ and $\tilde Q$ is required only when *estimating* our harm measures in a particular application, a topic we return to in the methods section.

@@

<!-- Despite being only a scalar quantity, $H(q)$ actually fully captures the harm created by the plan $q$.
We can zoom in to specific dimensions of the plan by conditioning on values of $V$, $\tilde Q$, and $I$, or by taking expectations.
Thus, for example, the harm for a single individual $i$ becomes $H(q)\mid I=i$, and the electorate's average harm in a particular election $v$ becomes
$$
    \E[H(q)\mid V=v] = \Pr[v(I)\neq\win_I(v, q) \text{ and }v(I)=\win_I(v, \tilde Q)],
$$
that is, the probability that a randomly selected voter in this election is misrepresented under the adopted plan but would not have been under a randomly selected counterfactual. -->

## Differential Harm {#sec-diff-harm}

@@

Once we have defined a set of groups, we define the *differential harm* for two groups $G_k$ and $G_l$ as
$$
  \Delta_{G_k-G_l}H(q) \dfeq \E[H(q)\mid I\in G_k] - \E[H(q)\mid I\in G_l].
$$
So $\Delta_{G_k-G_l}H(q)$ will be positive if voters in group $G_k$ are harmed at a higher rate than voters in group $G_l$, indicating that plan $q$ favors $G_l$ over $G_k$.

Zero differential harm between groups is analogous to the notion of *error rate balance* in the algorithmic fairness literature, which seeks to equalize false positive and false negative rates across protected classes [@zafar2017fairness; @chouldechova2017fair].
There, the implicit counterfactual is a perfect predictor.
Compared to the perfect predictor, the actual decision process harms individuals whenever it makes a false positive or false negative error.
Here, there can be no perfect districting plan, and the counterfactual instead consists of a collection of plausible alternatives.

@@

Estimating these quantities from available data is another matter, of course, and requires additional assumptions, as we describe in the methods section.

@@

To demonstrate the proposed framework in a classic partisan gerrymandering context, we provide a brief analysis of congressional districting in the state of New Jersey.
We show that differential harm correctly identifies possible gerrymandering, in contrast to some common redistricting metrics.
Appendices [-@sec-app-nj] and [-@sec-app-corr] contain further comparisons between gerrymandering metrics.

Since 1995, New Jersey congressional redistricting has been handled by a 13-member commission, with 6 political appointees from each party and one independent chair with *de facto* tiebreaking authority.
@fig-nj-maps shows the baseline Democratic vote share by precinct, along with the Democratic and Republican commission proposals.[^8]
In a typical election, the Democratic proposal yields 9 Democratic and 3 Republican seats, while the Republican proposal gives a 7--5 split.

[^8]: The Democratic plan was enacted. The GOP plan was provided by Harrison Neely, executive director for the GOP delegation to the New Jersey Congressional Redistricting Committee, following an Open Public Records Act request.

```{r fig-nj-maps, fig.width=6.25}
#| fig.cap: |
#|      Baseline vote patterns (a), two hypothetical gerrymanders (b) and (c),
#|      and one randomly sampled counterfactual plan (d).
#|      County lines are overlaid in white.
include_graphics("figures/nj_maps.pdf")
```

To provide a neutral comparison baseline, we generate 10,000 districting plans across using the SMC algorithm of [@mccartan2020; implemented in @redist] setting constraints to favor compact plans that respect county lines.
@fig-nj-maps(d) shows one such plan.
For each of these sampled plans and the two proposals, we estimate the average harm and differential partisan harm versus the simulated baseline,[^9] and calculate the expected number of Democratic seats and several common partisan metrics.
These measures, and their distribution in the basline, are shown in @fig-nj-meas.

[^9]: We choose to calculate $\Delta_{\dem-\rep}H(q)$, so positive values correspond to a Republican-favoring plan.

The estimate of differential partisan harm aligns with the expected partisan skew of each plan, with a value of $-3.6\%$ for the Democratic proposal and $2.0\%$ for the Republican proposal.
This means that, for example, an average Democratic voter is $3.6$pp less likely to be misrepresented than their average Republican counter-partisan.
Overall, the Democratic plan has lower average harm.

@@

Most of the other metrics, with the notable exception of the mean-median measure, agree with differential harm that each party's proposal favors itself.
The efficiency gap and partisan bias estimate give roughly equal scores in absolute value to the Democratic and Republican plans, in comparison to differential harm and declination, which identify the Democratic proposal as slightly more extreme.

The raw metrics can be better contextualized in comparison to the distribution across the sampled plans, something we discuss in [Appendix @sec-app-nj].
That appendix also compares the correlations between the various metrics in the simulated sample.
We find strong correlations of differential harm with the number of seats won by each party, while several common metrics like the mean-median gap and partisan bias are very weakly correlated with seats won.

@@

We next apply our proposed framework to @2022merrill, ongoing VRA litigation in which civil rights groups have argued that the existing Black majority district packs Black voters, and so a second Black opportunity-to-elect district should be drawn.
Measuring differential harm by both race and party will allow us to cleanly separate the racial and partisan effects of the Alabama districting plan, something existing approaches are unable to do.

Plaintiffs had to satisfy the first *Gingles* criterion by submitting one or more remedial maps demonstrating that a new Black opportunity-to-elect district could be drawn while meeting all other redistricting requirements.
In *Allen*, the plaintiff's expert, Moon Duchin, submitted four such plans, one of which is shown in @fig-al-maps(e).
These plans form a natural counterfactual $\tilde Q$ to use in analyzing the challenged districting plan (@fig-al-maps(d)), especially since the district court ruled in favor of the plaintiffs and found that their remedial plans satisfied the first *Gingles* criterion.
In [Appendix @sec-app-alsampling], we repeat the analysis performed here on a different counterfactual consisting of plans sampled with a redistricting simulation algorithm; the results are broadly similar.

@@

This analysis is slightly more complicated than in New Jersey, since in order to understand the racial distribution of harm we need to estimate turnout and voting preferences by race.
As we describe in the methods section, this is accomplished with an ecological inference model.
The estimates confirm that voting is highly polarized by race outside of the cities of Birmingham and Montgomery.

With these estimates in hand, @fig-al-maps(c) shows the average number of voters harmed in each precinct, and @fig-al-maps(f) summarizes harm by party and race.
Harm is concentrated among Black and Democratic voters in Birmingham, Montgomery, and the southern end of the Black Belt region.
These areas are excluded from the Black-majority district under the challenged plan, but are included in a Black-majority district in the average remedial plan.

Across the state, the estimated average harm is $7.9\%$.
The expected harm for a Black Democrat is $22\%$, for a Black Republican is $3.7\%$, for a White Democrat is $8.3\%$, and for a White Republican is $1.7\%$.
Thus while the challenged plan differentially harms both Black voters and Democratic voters, it *especially* harms Black Democrats, who have an estimated differential harm of $13.8\%$ versus White Democrats and of $18.3\%$ versus Black Republicans.

These calculations are made by aggregating harm within racial groups at the precinct level, and so they automatically take into account individual voter preferences.
Unlike in a traditional *Gingles* analysis, there is no need to identify a particular party as the uniform statewide "choice" of the minority racial group, nor is there any cause for concern because of lower racial voting polarization in urban areas.
The harm measure also treats voters of every race and location equally, recording just their inability to elect their preferred candidate.
<!-- Only once we have aggregated these individual harms does it become apparent that the challenged plan concentrates the harm among certain groups of voters. -->
